---
title: "Variance-Guided Time-Series Modeling with varGuidTS"
author: "Zihao Wang and Min Lu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Variance-Guided Time-Series Modeling with varGuidTS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

`varGuidTS` implements variance-guided penalized panel time-series models for
personalized temporal risk assessment. The package combines an autoregressive
exogenous mean model with a GARCH-X variance model, allowing both the expected
level and the variability of a repeated signal to depend on covariates.

The workflow is especially useful for wearable-sensor panels, where multimodal
physiological streams often show subject-specific baselines, short-memory
dynamics, motion artifacts, and time-varying volatility. The same modeling
structure can also be used for other balanced panel time series with repeated
measurements across individuals, devices, sites, or experimental units.

The package estimates:

- conditional mean dynamics with autoregressive and distributed-lag covariate effects;
- conditional variance dynamics with ARCH/GARCH terms and optional covariate effects;
- subject-specific baseline terms for both the mean and variance equations;
- sparse shared covariate effects through \(\ell_1\) penalization;
- exceedance-based risk scores, defined as conditional threshold-exceedance probabilities.

# Data layout

The input data frame must contain at least three columns:

| Column | Meaning |
|---|---|
| `s` | subject, panel, or unit identifier |
| `t` | integer time index within subject/unit |
| `y` | response or signal of interest |

All remaining columns are treated as exogenous predictors, except oracle columns
created by `simulate_scenario()` such as `mu`, `sigma`, `pi_true`, and
`scenario`, which are ignored by `lmvt()`.

# Model specification

Let \(y_{s,t}\) be the response for subject or unit \(s\) at time \(t\), and
let \(x_{s,t}\) be a vector of exogenous predictors. The model decomposes the
signal as

\[
y_{s,t} = \mu_{s,t} + \sigma_{s,t}\varepsilon_{s,t},
\]

where \(E(\varepsilon_{s,t} \mid x_{s,t}, \mathcal F_{t-1}) = 0\) and
\(\mathrm{Var}(\varepsilon_{s,t} \mid x_{s,t}, \mathcal F_{t-1}) = 1\). Thus
\(\mu_{s,t}\) is the conditional mean and \(\sigma_{s,t}\) is the conditional
standard deviation.

## Model orders

The function `lmvt()` specifies the model using four order parameters:

| Argument | Description |
|---|---|
| `p` | autoregressive order in the mean, using lags of `y` |
| `q` | distributed-lag order for exogenous predictors `x` |
| `r` | ARCH order in the variance, using lags of squared residuals |
| `s_ord` | GARCH order in the variance, using lags of conditional variance |

When `use_x_in_variance = TRUE`, the same lag block of predictors
\((x_{s,t}, x_{s,t-1}, \ldots, x_{s,t-q})\) enters both the mean and variance
equations. This choice keeps the model parsimonious and stable when the number
of time points per unit is modest.

## Mean equation

The ARX mean model is

\[
\mu_{s,t} = \alpha_s + \sum_{i=1}^{p}\theta_i y_{s,t-i}
+ \sum_{j=0}^{q} x_{s,t-j}^{\top}\beta_j.
\]

The subject-specific intercept \(\alpha_s\) captures baseline differences, while
\(\theta_i\) and \(\beta_j\) are shared population-level effects. The covariate
coefficients \(\beta_j\) can be estimated under an \(\ell_1\) penalty controlled
by `lambda_beta`.

## Variance equation

The GARCH-X variance model is

\[
\sigma_{s,t}^2 = \omega_s
+ \sum_{\ell=1}^{s_{\mathrm{ord}}} b_\ell \sigma_{s,t-\ell}^2
+ \sum_{h=1}^{r} a_h e_{s,t-h}^2
+ \sum_{j=0}^{q} x_{s,t-j}^{\top}\gamma_j,
\]

where \(e_{s,t}=y_{s,t}-\mu_{s,t}\). The baseline variance \(\omega_s\) is
subject-specific, while the ARCH parameters \(a_h\), GARCH parameters
\(b_\ell\), and variance covariate effects \(\gamma_j\) are shared across units.
The variance covariate effects can be penalized through `lambda_gamma`.

The implementation enforces positive variance through non-negativity constraints
and prediction-time feasibility repair.

# Basic workflow

Prepare a balanced panel with columns for subject ID `s`, time index `t`, response `y`, and covariates.
Fit the penalized ARX--GARCHX model using `lmvt()`.
Use `summary()` to inspect selected mean and variance covariates.
Use `predict()` to compute conditional means, variances, and exceedance-based risk scores.

```{r basic-workflow, eval=FALSE}
library(varGuidTS)

set.seed(1)
sim <- simulate_scenario(scen = 2, S = 4, T = 60,
                         d_noise = 4, noise_kind = "ar1", seed = 1)

df <- sim[, c("s", "t", "y",
              grep("^X|^Noise|^Xbin", names(sim), value = TRUE))]

fit <- lmvt(df, p = 1, q = 0, r = 1, s_ord = 1,
            lambda_beta = 0.05, lambda_gamma = 0.05,
            use_x_in_variance = TRUE, maxit = 8)

fit
```

# Coefficient summaries

The `summary()` method returns model-level information and a coefficient table containing coefficient block, equation, term name, estimate, absolute estimate, subject identifier for subject-specific terms, and a selected/nonzero indicator.

```{r coefficient-summaries, eval=FALSE}
sm <- summary(fit, include_zero = FALSE, sort_by = "abs_estimate")
sm$model
sm$nonzero
head(sm$table)
```

# Prediction and exceedance-based risk scores

The `predict()` method returns estimated conditional means and variances. When a threshold is supplied, it also returns exceedance probabilities. A single pooled threshold may be used, or a vector of subject-specific thresholds may be supplied with names matching the subject IDs.

```{r prediction, eval=FALSE}
# Pooled threshold
pred_global <- predict(fit, df, threshold = quantile(df$y, 0.90))
head(pred_global)

# Subject-specific thresholds
subject_cutoffs <- tapply(df$y, df$s, quantile, probs = 0.75, na.rm = TRUE)

pred_subject <- predict(fit, df,
                        threshold = subject_cutoffs,
                        innov_g = TRUE,
                        innov_t = TRUE,
                        df_t = 8)
head(pred_subject)
```

Here `df_t = 8` specifies 8 degrees of freedom for the standardized Student-t innovation distribution used to compute `risk_t`; it is not related to the example data frame named `df`.

Mathematically, for a threshold \(c\), the score is

\[
\hat\pi_{s,t}(c) = P(Y_{s,t} > c \mid \mathcal F_{t-1}, x_{s,t})
= 1 - F_\varepsilon\left(\frac{c - \hat\mu_{s,t}}{\hat\sigma_{s,t}}\right).
\]

# Simulation scenarios
`simulate_scenario()` provides four canonical scenarios for examples and method
checks:

| Scenario | Description |
|---|---|
| `scen = 1` | baseline setting with stable variance |
| `scen = 2` | mean-driven drift or seasonality |
| `scen = 3` | short regime-driven variance bursts |
| `scen = 4` | seasonal structure with piecewise variance regimes and heavy-tailed innovations |

```{r scenarios, eval=FALSE}
s1 <- simulate_scenario(scen = 1, S = 3, T = 50, d_noise = 3, seed = 10)
s2 <- simulate_scenario(scen = 2, S = 3, T = 50, d_noise = 3, seed = 20)
s3 <- simulate_scenario(scen = 3, S = 3, T = 50, d_noise = 3, seed = 30)
s4 <- simulate_scenario(scen = 4, S = 3, T = 50, d_noise = 3, seed = 40)
```

The simulation output contains oracle columns such as `mu`, `sigma`, and
`pi_true`, which are useful for method checks but are ignored when fitting the
model.

# Interpretation

The exceedance-based risk score is a statistical summary of conditional
threshold-exceedance behavior. It can be used as an interpretable temporal
feature or as an input to downstream modeling. It should not be interpreted as a
clinically validated diagnostic probability unless validated in an appropriate
prospective study.
